Appendix 3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB

### 2021-12-22 doc/appendix-snow-sdmTMB.Rmd
Make snowy app. run

```{r zero-test, echo=TRUE, message=FALSE, cache=FALSE}
s_nb2 <- simulate(m, nsim = 500)
pred_fixed <- m$family$linkinv(predict(m)$est_non_rf)
r_nb2 <- DHARMa::createDHARMa(
  simulatedResponse = s_nb2,
  observedResponse = m$data$count,
  fittedPredictedResponse = pred_fixed
)
DHARMa::testZeroInflation(r_nb2, plot = FALSE)
```

@@

```{r plot-resids, fig.width = 8, cache=FALSE}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.5) +
  facet_wrap(~year) +
  coord_fixed() +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

@@

We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from [http://www.naturalearthdata.com/](http//www.naturalearthdata.com/download/10m/physical/ne_10m_lakes.zip) and transformed them to the correct projection.

```{r shapes}
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers) # only large lakes
```

@@

```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=FALSE}
b <- tidy(m)
p_mean <- p %>%
  group_by(CBCID) %>%
  summarise(
    X = mean(X), Y = mean(Y),
    mean_est_count = exp(mean(est)),
    nao_effect = exp(b[b$term == "nao", 2] + mean(zeta_s))
  )
```

```{r plot-map, echo=TRUE, eval=T, cache=T}
ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  guides(size = "none") +
  labs(colour = "NAO effect\non Snowy Owl count") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

@@

```{r plot-all-effects, echo=TRUE, eval=T, cache=T}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatial-effects, echo=TRUE, eval=T, cache=T}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

@@

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=T}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-fixed-effects, echo=TRUE, eval=T, cache=T}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

```{r zero-test, echo=TRUE, message=FALSE}
s_nb2 <- simulate(m, nsim = 500)
pred_fixed <- m$family$linkinv(predict(m)$est_non_rf)
r_nb2 <- DHARMa::createDHARMa(
  simulatedResponse = s_nb2,
  observedResponse = m$data$count,
  fittedPredictedResponse = pred_fixed
)
DHARMa::testZeroInflation(r_nb2, plot = FALSE)
```

@@

```{r plot-resids, fig.width = 8}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.5) +
  facet_wrap(~year) +
  coord_fixed() +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

@@

We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from <http://www.naturalearthdata.com/> and transformed them to the correct projection.

```{r shapes}
if (!file.exists(here::here("data/ne_10m_lakes"))) {
  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
    "download/10m/physical/ne_10m_lakes.zip")
  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
}
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers) # only large lakes
```

@@

```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE}
b <- tidy(m)
p_mean <- p %>%
  group_by(CBCID) %>%
  summarise(
    X = mean(X), Y = mean(Y),
    mean_est_count = exp(mean(est)),
    nao_effect = exp(b[b$term == "nao", 2] + mean(zeta_s))
  )
```

```{r plot-map, echo=TRUE, eval=TRUE}
ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  guides(size = "none") +
  labs(colour = "NAO effect\non Snowy Owl count") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

@@

```{r plot-all-effects, echo=TRUE, eval=TRUE}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatial-effects, echo=TRUE, eval=TRUE}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

@@

```{r plot-spatiotemporal-effects, echo=TRUE, eval=TRUE}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-fixed-effects, echo=TRUE, eval=TRUE}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```
